---
title: 'Tutorial: Downloading data'
output:
  html_document: default
  word_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```
\
\

### Timings 
  1. Download bat data (ca. 35 min)
  2. un-bzip the files and gzip them so they are ready for Stacks 2 ( ca. 5 min)
  
All done in EVE cluster
  
\
\

### 1. Download ENA data and get them in correct format for Stacks 2

This is all done in the EVE cluster, you can make your own scripts to do this or use the one provided #(run_Admixture.sh). It can be submitted via sbatch in the EVE cluster, but be sure to modify the email address #and output paths to your own details.


```{r}
# download ENA Data


cd /data/life_on_the_edge/
mkdir Workshop_3/
mkdir Workshop_3/Myotis_escalarai/
cd Workshop_3/Myotis_escalarai/
mkdir raw/
mkdir demultiplexed/
cd raw/

# bat data example download == 134 inds of Myotis bats ddRADseq PE data, sbfI and nlaIII enzymes. barcodes are inline inline adapters

#https://www.ebi.ac.uk/ena/browser/text-search?query=PRJEB29086
wget ftp://ftp.sra.ebi.ac.uk/vol1/run/ERR283/ERR2832479/BAT_01_NoIndex_L008_R1_001.fastq.bz2
wget ftp://ftp.sra.ebi.ac.uk/vol1/run/ERR283/ERR2832479/BAT_01_NoIndex_L008_R2_001.fastq.bz2

# these are bzip2ed, so need to be unzipped and then gzipped for Stacks

ml GCCcore/10.2.0
ml bzip2/1.0.8
cd /data/life_on_the_edge/Workshop_3/Myotis_escalarai/raw/

bzip2 -d ./*
gzip ./*

```{r}
