---
title: 'Tutorial: denovo_map'
output:
  html_document: default
  word_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```
\
\

### Timings 
  1. Run denovo_map (2 hours?)
  2. Identify  bad apples and rerun denovo_map (1 hour?)
  
All steps done in EVE cluster  
  
\
\

### Objectives

This exercise will show you how to run denovo_map from start to finish (assuming that you have demultiplexed data after running process_radtags). For this exercise we will use some already processed data for Hyphaene coriacea, a palm species from Madagascar

\
\

### 1. Run denovo_map

This is all done in the EVE cluster, you can make your own scripts to do this or use the one provided (run_Admixture.sh). It can be submitted via sbatch in the EVE cluster, but be sure to modify the email address and output paths to your own details

```{r}
LAURA's CLEANED CODE
```

\
\

### 2. Identify bad apples and rerun denovo_map
```{r}
LAURA's CLEANED CODE
```

