---
title: 'Tutorial: Discriminant Analysis of Principal Components (DAPC)'
output:
  html_document: 
      theme: readable
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, 
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
knitr::include_graphics
```

### Timings

  1. Initial setup and package install (<1 min)
  2. Load data abd convert to genind format for Adegenet (< 1 min)
  3. Run the DAPC and plot the BIC scores (3 mins)
  4. Make barplot of Admixture results (< 1 min)

All done in R locally on your machine

### Other info
- If you are on a PC, directory paths need to be separated by back-slashes instead of forward-slashes
- Try not to resize windows/display panels while R is busy calculating/plotting - this tends to cause problems

### 1. Initial setup and package install
The following block of code will install all required R packages for this tutorial, you can copy and paste this in your R Studio and run it. All of this code works with the latest R version (4.1.0), and possibly older ones (e.g. 3.5.0), but I cannot guarantee it will work, so best to have R v.4.1.0.
```{r}
rm(list=ls(all=TRUE))

#install.packages(adegenet)
library(adegenet)
```

### 2. Load data and convert to genind format for Adegenet
Load the .dat input file that we will analyse. I've already converted this for you into a .dat (fstat format) file. Easiest way to do this is using PGDSpider (http://www.cmpg.unibe.ch/software/PGDSpider/).

Again, this is a fancy file structure which is basically just a matrix of individuals and genotypes across all SNPs...

```{r}
# Read in .dat file (fstat) format needed by Adegenet package for running DAPC
data <- read.fstat('./data/L_flavomaculatus/Outputs/DAPC/Lflavomaculatus.dat')
```

We can check the size of the file with the length argument (should be 59 individuals)
```{r}
length(indNames(data))
```


### 3. Run the DAPC and plot BIC scores
We then run the DAPC itself using find.clusters, max.n.clusters can be up to any number (<indiviudals), but let's set it to 10 here to keep in line with the Admixture and sNMF analyses. Take all possible PCAs (so 60, and press [enter]
```{r}
grp <- find.clusters(data, max.n.clust=10, n.pca=59, choose.n.clust = FALSE, criterion = "min") 
# the last 3 arguments here have automated this, highly recommended to remove these and do it interactively in R by looking at the resulting plots
# Advice:
# number of PCs to retain - here say (number of individuals - 1 to maximise the information initially)
# number of clusters - here select a sensible lowest BIC. In our case we automatically select 4 (min) because the elbow is very clear, in other datasets this my not be the case...
```
It will ask you to choose the number of clusters. As you can see, K=4 again is best. We can look at some summary stuff here which is useful (BIC scores, best K, Population assignment per individual, number of Inds in each population cluster:
```{r}
head(grp$Kstat) # BIC scores per k
grp$stat # Best k based on BIC
head(grp$grp, 59) # Group membership (based on your defined number of clusters)
grp$size # Group size per k cluster

x <- as.data.frame(grp$Kstat)
png(filename="./outputs/DAPC_Results_BIC_scores.png", type="quartz", height=8, width=8, units="in", res=300)
plot(x$`grp$Kstat`, pch=19, col= "black", cex=0.25, xlab="k clusters", ylab="Bayesian Information Criterion", main = "BIC across k")
lines(x$`grp$Kstat`, lty=1, lwd=1.5, col='steelblue2')
dev.off()
```


Now let's do the DAPC. Run the below code. Thsi time choose a lower number of retained PCAs, ideally where the curve starts to become less (e.g. 4 or 5 in this case). Press [enter] and then choose the number of discriminant functions to retain based on the eigenvalues. In this case you want to take all 3, but depends on the information content (F-statistic).

```{r}
# now do the DAPC
dapc1 <- dapc(data, grp$grp, n.pca = 5, n.da = 3)
# same as above, this is usually interactive - the last 2 arguments here have automated this, highly recommended to remove these and do it interactively in R by looking at the resulting plots
# Advice:
# number of PCs to retain - now we can be more selective, pick a number where the initial curve of the plot starts to slow (in our case 4 or 5)
# number of discriminant functions to retain - always more than one - general rule is to take anything which is quite a lot higher than 0, so in this case say 3. If our eigenvalues here showed two big bars and then a very small one we would take 2. More details on how to interpret and select are found here: https://adegenet.r-forge.r-project.org/files/tutorial-dapc.pdf

```

### 4. Make DAPC plot of results (< 1 min)


Set some colours and then plot
```{r}
myCol <- c("blue","turquoise","lightblue", "darkblue")
png("./outputs/DAPC_Lflavomaculatus.png",width=500,height=500)
scatter(dapc1, scree.da=TRUE, scree.pca=TRUE, bg="white", pch=20, cell=0, cstar=0, col=myCol, solid=1, cex=1.5, clab=0, leg=FALSE, txt.leg=paste("Cluster",1:4))
dev.off()
```



### On your own...

Mess around with PCAs retained and number of clusters to see how it affects the BIC values, curves and population assignment. It can be really tricky without god knowledge of true K, especiallly in not well structured populations