---
title: 'Tutorial: snMF'
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```
\
\

### Timings 

  1. Load data (< 1 min)
  2. Run sNMF (3-4 mins)
  3. Plot CV scores (< 1 min)
  4. Make barplot of Admixture results (< 1 min)

All done in R locally on your machine
  
\
\

### Objectives

Again, using the same [Leptopelis flavomaculatus](https://amphibiaweb.org/species/3649) data we will try a slightly different but very similar approach to look at population structure. Instead of Admixture we will use [sNMF](http://membres-timc.imag.fr/Olivier.Francois/LEA/tutorial.htm) from the LEA R package this time

\
\

### 1. Load data
Load the .ped input file that we will analyse. This is a matrix (rows = number of individual samples, columns = genotypes at each locus). The first 6 columns are mandatory (family, individual, maternal/paternal IDs, Sex abd Phenotype). We actually only have the individual IDs so the others are defaulted out. You can read more about this PLNK, format here: https://zzz.bwh.harvard.edu/plink/data.shtml

```{r}
# Read in ped file, convert to .geno format needed by LEA package for sNMF
library(LEA)

data = read.table('/Users/cb76kecu/Desktop/ddRAD-seq_workshop/data/Exercises_4-8/sNMF/Lflav.ped')
output = ped2geno('/Users/cb76kecu/Desktop/ddRAD-seq_workshop/data/Exercises_4-8/sNMF/Lflav.ped')
geno <- read.geno('/Users/cb76kecu/Desktop/ddRAD-seq_workshop/data/Exercises_4-8/sNMF/Lflav.geno') 
```

\
\

### 2. Run sNMF
We can give sNMF a range of K to test, so let's follow 1:10 like we did in Admixture so that we have comparable results. Here we set the ploidy to 2 (diploid organism, having 5 repetitions of each K value to test, so 50 total reps). We set entropy=T and project='new'). There are a bunch of things you can tweak (see here: http://membres-timc.imag.fr/Olivier.Francois/snmf/files/note.pdf) but let's keep this simple. Current settings here will take around 3-4 mins to complete

Block to update:
```{r}
# Run snmf to look at structure 
k_range <- 1:10
geno.snmf <- snmf(('/Users/cb76kecu/Desktop/ddRAD-seq_workshop/data/Exercises_4-8/sNMF/Lflav.geno'), K = k_range, entropy = T, ploidy = 2, rep=5, project = 'new')  
```

\
\

### 3. Plot CV scores
We will first set the plot margins and then set an output file (png) with specific sizes and resolutions
Then we plot the results (first points, then a line to join the points) and save the file
```{r}
# plot CV scores
sumpr2 <- summary(geno.snmf)$crossEntropy
sumpr <- sumpr2[2,]

par(mar=c(2,2,2,2))
png(filename="/Users/cb76kecu/Desktop/ddRAD-seq_workshop/outputs/Exercise_5/sNMF_Results_CV_scores.png", type="quartz", height=8, width=8, units="in", res=300)
plot(sumpr, lty=2, lwd = 1.5, col = 'black', pch=19, cex=0.5, xlab= 'k clusters', ylab = 'CV score', main= 'Cross validation scores across k (sNMF)')
lines(sumpr, lwd=2, col='steelblue2')
dev.off()
```

\
\

### 4. Make barplot of sNMF results

We'll first do an automated thing to get the best value of K (which K value and which run of the 5 you made)
```{r}
# Choose K with lowest mean cross entropy
K <- c(1:10)[which.min(sumpr2[2,])]         # K=4
# Get the cross-entropy of each run for K
ce <- cross.entropy(geno.snmf, K=K)
# Select the run with the lowest cross-entropy
best <- which.min(ce)
best
```

Next we will take the q values from the sNMF results for the selected K value and run (analagous to the Admixture Lflav.4.Q file). It is a matrix of 59 rows (one per individual) and 4 columns (i.e. K). The probability of membership to each population cluster is shown for each individual
```{r}
Q_vals <- Q(geno.snmf, K = K, run = best)
sample_names <- data[,2]
pop_assignment <- as.data.frame(cbind(sample_names,Q_vals))

ord = pop_assignment[order(pop_assignment$V4,pop_assignment$V1,pop_assignment$V2,pop_assignment$V3),]
rownames(ord) <- ord$sample_names
ord <- ord[,2:5]

par(mar=c(10,4,1,1))
bp = barplot(t(as.matrix(ord)), space = (0.04), col=c("blue", "turquoise", "darkblue", "lightblue"), xlab="", ylab="Ancestry", border=1, las=2, cex.names=0.9)
```

The above plot is not so clean, very messy and not very pretty. Let's reorder the output file so that it matches the order of individuals in the Admixture analyses. It can then be plotted for direct and easy comparison.
```{r}
# match ord_labels to the pop_assignment
ord_labels <- read.csv('/Users/cb76kecu/Desktop/ddRAD-seq_workshop/data/Exercises_4-8/sNMF/ord_labels_sNMF.csv')
ord_labels_sample_name_only <- as.data.frame(ord_labels);colnames(ord_labels_sample_name_only) <- 'sample_name_only'
ord_labels_sample_name_only <- substr(ord_labels_sample_name_only$sample_name_only, 1, 5)
ord_labels_sample_name_only <- as.data.frame(ord_labels_sample_name_only)
ord_labels <- cbind(ord_labels,ord_labels_sample_name_only)

# now sort pop_assignment based on order in ord_labels$sample_name_only

sorted_pop_assignment <- pop_assignment[order(match(pop_assignment[,1],ord_labels_sample_name_only[,1])),]
row.names(sorted_pop_assignment) <- sorted_pop_assignment$sample_names
sorted_pop_assignment <- sorted_pop_assignment[,2:5]

# plot the sNMF barplot for comparison with Admixture
bp = barplot(t(as.matrix(sorted_pop_assignment)), space = (0.04), col=c("blue", "lightblue", "turquoise", "darkblue"), xlab="", ylab="Ancestry", border=1, las=2, cex.names=0.9)

# save
png(filename="/Users/cb76kecu/Desktop/ddRAD-seq_workshop/outputs/Exercise_5/sNMF_Results_Barplot.png", type="quartz", height=8, width=30, units="in", res=300)
bp = barplot(t(as.matrix(sorted_pop_assignment)), space = (0.04), col=c("blue", "lightblue", "turquoise", "darkblue"), xlab="", ylab="Ancestry", border=1, las=2, cex.names=0.9)
dev.off()

```

\
\

### On your own...

If you want, try plotting k=1 - k=5, feel free to change colour schemes and the order of the tbl object
If you have your own data you can try and use this script to plot your own outputs
