---
title: 'Tutorial: RAxML'
output:
  html_document: default
  word_document: default
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, eval = FALSE,
                      echo=TRUE, warning=FALSE, message=FALSE,
                      tidy = TRUE, collapse = TRUE,
                      results = 'hold')
```
\
\

### Timings 
  1. Run RAxML (ca. 5 min for 5 bootstraps and full MLE search, higher numbers of bootstraps will run increasingly longer)
  
All done in EVE cluster
  
\
\

### Objectives

We've looked at 3 methods for looking at population structure. We can also confirm this information on population structure from a phylogeny, by looking at how the individuals and populations cluster within a tree. More interestingly we can look at the relationships between populations (something that is not particularly clear with Admixture, sNMF and DAPC analyses)

\ 

For this exercise we will use [RAxML](https://cme.h-its.org/exelixis/web/software/raxml/) on the EVE cluster to investigate phylogenetic relationships. Note that we will not do any phylogenetic dating here, for that you could use a variety of other things, for example [BEAST](https://beast.community/). 

\
\

### 1. Run RAxML
#This is all done in the EVE cluster, it will take some time - we can reduce the number of bootstrap replicates to 5. You can build your own script or use the supplied one that you can edit (07_RAxML.sh)

```{r}
#!/bin/bash                                                                                                                            

#SBATCH --job-name=RAxML
#SBATCH --mail-user=christopher_david.barratt@uni-leipzig.de
#SBATCH --mail-type=BEGIN,END,FAIL,TIME_LIMIT
#SBATCH --output=/work/$USER/ddRAD-seq_workshop/job_logs/%x-%j.log
#SBATCH --cpus-per-task=1 
#SBATCH --mem-per-cpu=40G
#SBATCH --time=12:00:00

#load environment
module purge
module load Anaconda3
source activate /global/apps/raxml_8.2.9

cd /work/$USER/ddRAD-seq_workshop/data/Exercises_4-8/RAxML/

# 10 bootstraps


raxmlHPC -f a -m GTRCAT -p 12345 -x 12345 -# 5 -s  /work/$USER/ddRAD-seq_workshop/data/Exercises_4-8/RAxML/Lflav.phy ­­asc­corr=lewis -n bootstrapped 
raxmlHPC -m GTRCAT -p 12345 -f b -t RAxML_bestTree.bootstrapped -z RAxML_bootstrap.bootstrapped -n FINAL_bootstrapped.tre

# move output files to outputs directory
mv /work/$USER/ddRAD-seq_workshop/data/Exercises_4-8/RAxML/RAxML*.* /work/$USER/ddRAD-seq_workshop/outputs/Exercise_7/
```

\
\



### Visualise the tree

First we can run rsync again to pull in the files from RAxML into our desktop directory:

\

```{r}
# transfer from the cluster
rsync -avhP -e 'ssh -p 8022' \
  barratt@localhost:/work/$USER/ddRAD-seq_workshop/outputs/ \
  /Users/cb76kecu/Desktop
```

\

Then...

* Download and install [Figtree](https://www.softpedia.com/dyn-postdownload.php/e272921fefc6c686b1fa2b06df64a860/62a88e68/2a33f/4/1) on your computer. 

* Rsync the files from the EVE cluster back to your own computer again

* Open Figtree and open the "RAxML_bipartitions.FINAL_bootstrapped.tre" file

* A window will come up, just click ok (it's about how to visualise the bootstrap values on nodes)

* Your tree will appear, let's visualise it better...

* Open the "Trees" panel on the left, tick the "Root tree" box and then select "Rooting: Midpoint"

* Go to the main menu and say "File > Import Annotations". When the dialog box opens select the ```ddRAD-seq_workshop/data/Exercises_4-8/RAxML/RAD_IDs.txt``` file

* Go back to Figtree and then open the "Tip Labels" panel, select "Display: NEW NAME"

* You'll see that the tip labels have now been replaced and you can compare them easily with the Admixture, sNMF and DAPC results

\
\

### On your own...

You can rerun RAxML with different parameter settings, for example by changing the number of bootstraps by changing the number in the first line to -# 10 instead of -# 5 (100 recommended for a full analysis, but this takes time). You could also look at different models of evolution instead of GTRCAT, or try a different ascertainment bias correction (not lewis)
